---
title: "Cancer EDA"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
```
# Introduction

Research Question: Perform an exploratory analysis to understand how county-level characteristics are related to cancer mortality.

```{r}
setwd("~/Desktop/Courses/W203 - Stat for Data Science/Lab_1/MIDS_LAB01")
data = read.csv("cancer.csv")
```

Number of Variables: `r length(colnames(data))`  
Number of Observations: `r nrow(data)`  

### Variables: 
This dataset contains variables describing county, region, population, birthrate, race, marital status, insurance coverage, income status, and education.
```{r}
colnames(data)
```

## Variable clarification and assumption
PctPrivateCoverage: "Percentage of the population with private insurance coverage"  
       avgAnnCount: "2009-2013 mean incidences per county WHAT DOES THIS MEAN????"  
    povertyPercent: "Percent of population below poverty line"  
        popEst2015: "Estimated population by county 2015"  
 PctPublicCoverage: "Percentage of the population with public insurance coverage"  
         deathRate: "Number of deaths attributed to cancer"  
         binnedInc: "Income groups????"
         medianAge: "We removed all median ages above 100"


## Data Quality
1) The sample size seems to be large enough to get valuable insight.
2) The data seems to be collected in different number formats, even for the same columns. Some have integers, some have floats with one decimal, others many decimals.
3) Seems to be a number of obeservations that are NA of 18-24 with some college, `r sum(is.na(data$PctSomeCol18_24))` to be exact.

# Univariate Analysis of Key Variables

The key variables that we focused on are in groups related to the variable deathRate:

### Death Rate
```{r}
summary(data$deathRate)
deathPercentOfPopulation = data$deathRate/data$popEst2015
hist(data$deathRate, main = "Death Rate Frequency", xlab = "Deaths Per County")
```
The avg deathrate of cancer is between 150-200

### Population: popEst2015, AvgHouseholdSize, PercentMarried, Geography, avgAnnCount, BirthRate, binnedInc

```{r}
plot(data$popEst2015 ~ data$BirthRate)
# Higher population does not correlate with higher birth rate
plot(
  data$deathRate ~ data$popEst2015,
  main = "Death rate and Population",
  xlab = "Population",
  ylab ="Death Rate of Pop"
)
# Higher population does not correlate with high cancer deathrate
plot(
  deathPercentOfPopulation ~ data$BirthRate, 
  main = "Death rate and birth rate percent of Pop.",
  xlab = "Birth Percent of Pop.",
  ylab ="Death Rate Percent of Pop"
)
```
Conclusion on Population variables: 
- It doesn't seem that high birth rate or population correlates to higher cancer mortality.
- We removed avg household size, percent married, geography, angAnnCount, and binned income from analysis due to perceieved irrelevance.

### Age: MedianAge, MedianAgeMale, MedianAgeFemale

```{r}
reasonable_median_age = subset(data, MedianAge <= 100)
plot(
  reasonable_median_age$deathRate ~ reasonable_median_age$MedianAge,
  main = "Median Age And Cancer Death Rate",
  xlab = "Median Age",
  ylab = "Death Rate"
)

plot(
  reasonable_median_age$deathRate ~ reasonable_median_age$MedianAgeMale,
  main = "Median Age Male And Cancer Death Rate",
  xlab = "Median Age Male",
  ylab = "Death Rate"
)

plot(
  reasonable_median_age$deathRate ~ reasonable_median_age$MedianAgeFemale,
  main = "Median Age Female And Cancer Death Rate",
  xlab = "Median Age Female",
  ylab = "Death Rate"
)
```
Conclusion of Age Variables:
- We removed all median ages above 100 due to some anomalies of median age 300+.
- There seems to be a large cancer mortality rate between the 30-50 years of age.
- Women seem to group just above 40 and men just under 40 with county deathrates.

### Income: medIncome, povertyPercent, binnedInc, PctEmployed16_Over, PctUnemployed16_Over

```{r}
plot(
  data$deathRate ~ data$medIncome,
  main = "Median Income And Cancer Mortality",
  xlab = "Median Income",
  ylab = "Death Rate"
)
plot(
  data$deathRate ~ data$PctEmployed16_Over,
  main = "Employed 16yrs old or older And Cancer Mortality",
  xlab = "People Employed 16 Years Old or Older",
  ylab = "Death Rate"
)
plot(
  data$deathRate ~ data$povertyPercent,
  main = "People In Poverty And Cancer Mortality",
  xlab = "People In Poverty",
  ylab = "Death Rate"
)
plot(
  data$deathRate ~ data$PctUnemployed16_Over,
  main = "Unemployed 16yrs old or older And Cancer Mortality",
  xlab = "People Unemployed 16 Years Old or Older",
  ylab = "Death Rate"
)
```

Conclusion on Income Variables:
- The strongest correlation yet
- As poverty and unemployment goes up, so does cancer mortality
- As median income and employment rise, cancer mortality decreases

### Race: PctWhite, PctBlack, PctAsian, PctOtherRace

```{r}
plot(
  data$deathRate ~ data$PctWhite,
  main = "Percent Race 'White' And Cancer Mortality",
  xlab = "Percent Race 'White'",
  ylab = "Death Rate"
)
plot(
  data$deathRate ~ data$PctBlack,
  main = "Percent Race 'Black' And Cancer Mortality",
  xlab = "Percent Race 'Black'",
  ylab = "Death Rate"
)
plot(
  data$deathRate ~ data$PctAsian,
  main = "Percent Race 'Asian' And Cancer Mortality",
  xlab = "Percent Race 'Asian'",
  ylab = "Death Rate"
)
plot(
  data$deathRate ~ data$PctOtherRace,
  main = "Percent Race 'Other' And Cancer Mortality",
  xlab = "Percent Race 'Other'",
  ylab = "Death Rate"
)
```

Conclusion on Race Variables:
- It seems that many of the counties surveyed were a moajority race 'White'
- The death rate seemed to hover around its avg for every race, no major correlation detected

### Education: PctNoHS18_24, PctHS18_24, PctHS25_Over, PctSomeCol18_24, PctBachDeg18_24, PctBachDeg25_Over

```{r}
plot(
  data$deathRate ~ data$PctNoHS18_24,
  main = "Percent No HS Education 18-24 And Cancer Mortality",
  xlab = "Percent No HS Education 18-24",
  ylab = "Death Rate"
)
plot(
  data$deathRate ~ data$PctHS18_24,
  main = "Percent HS Education 18-24 And Cancer Mortality",
  xlab = "Percent HS Education 18-24",
  ylab = "Death Rate"
)
plot(
  data$deathRate ~ data$PctHS25_Over,
  main = "Percent HS Education 25+ And Cancer Mortality",
  xlab = "Percent HS Education 25+",
  ylab = "Death Rate"
)
plot(
  data$deathRate ~ data$PctSomeCol18_24,
  main = "Percent Some College 18-24 And Cancer Mortality",
  xlab = "Percent Some College 18-24",
  ylab = "Death Rate"
)
plot(
  data$deathRate ~ data$PctBachDeg18_24,
  main = "Percent College Grad 18-24 And Cancer Mortality",
  xlab = "Percent College Grad 18-24",
  ylab = "Death Rate"
)
plot(
  data$deathRate ~ data$PctBachDeg25_Over,
  main = "Percent College Grad 25+ And Cancer Mortality",
  xlab = "Percent College Grad 25+",
  ylab = "Death Rate"
)
```
Conclusion of Education Variables:
- As a population is more educated, cancer mortality falls
- It seems that college grauates make up less percent of cancer moratility population


### Insurance Coverage: PctPrivateCoverage, PctEmpPrivCoverage, PctPublicCoverage

```{r}
plot(
  data$deathRate ~ data$PctPrivateCoverage,
  main = "Percent of Private Insurance And Cancer Mortality",
  xlab = "Percent of Private Insurance",
  ylab = "Death Rate"
)
plot(
  data$deathRate ~ data$PctEmpPrivCoverage,
  main = "Percent of Employer Private Insurance And Cancer Mortality",
  xlab = "Percent of Employer Private Insurance",
  ylab = "Death Rate"
)
plot(
  data$deathRate ~ data$PctEmpPrivCoverage,
  main = "Percent of Public Insurance And Cancer Mortality",
  xlab = "Percent of Public Insurance",
  ylab = "Death Rate"
)
```

Conclusion on Insurance Variables:
- It seems when a population has private insurance cancer mortality is down
- When a population has public insurance, cancer mortality is up
- Public coverage could be correlated by income/poverty/unemployment




















