---
title: "Cancer Mortality Exploration"
author: "w203 Teaching Team"
output: pdf_document
---

## Background

In this lab, imagine that your team is hired by a health government agency. They would like to understand factors that predict cancer mortality rates, with the ultimate aim of identifying communities for social interventions, and of understanding which interventions are likely to have the most impact.  Your team was hired to perform an exploratory analysis to help the agency address their goals.


## Data

You are given a dataset for a selection of US counties, **"cancer.csv"**.  The dependent (or target) variable in this data is named "deathRate".

The labels of some of the variables are listed below; the rest of the variables should be self-explanatory.

```{r}
# Load CSV data
cancer = read.csv("cancer.csv")
```

## Introduction

```{r}
# Show variable names
colnames(cancer)
```

```{r}
# Show summary
summary(cancer)
```

```{r}
# Show summary
length(cancer)
```

```{r}
# Show how many observations there are in each variable
for(df in cancer) {
  cat(sprintf("Dataframe length: %d\n", length(df)))
}
```

## Univariate Analysis of Key Variables

```{r}
# Create a working copy of the variable cancer
canc <- cancer

# Summary
summary(canc$deathRate)

# Histogram
hist(canc$deathRate, breaks = 20, main = "Death Rate", xlab = NULL)
```

```{r}
# Define a function for running a set of compareative methods such as cor(), scattered plot, etc
compare_with_deathRate <- function(c, df, count) {
  if(is.na(colnames(c)[count])) {
    ylabel =  NULL
  } else {
    ylabel = colnames(c)[count]
  }

  # DEBUG  
  #cat(sprintf("Dataframe name: %s\n", colnames(c)[df]))
  
  # Correlation
  cat(sprintf("Correlation - Death Rate and %s: %f\n", ylabel, cor(c$deathRate, df)))
  
  # Scatter Plot
  plot(jitter(c$deathRate, factor=2), jitter(df, factor=2),  
       xlab = "Death Rate", ylab = ylabel, 
       main = "Correlation with Death Rate")
  abline(lm(c$deathRate ~ df))
}

# Loop over all numeric variables in canc and analyze
count <- 1
cors = vector('numeric')

for(df in canc) {
  if(class(df) == "numeric" && colnames(canc)[count] != "deathRate") {
    compare_with_deathRate(canc, df, count)
    cors[count] = cor(canc$deathRate, df)
  }
  count = count + 1
}

tail(sort(cors),5)
head(sort(cors),5)
```

### Potentially erroneous data

```{r}
# Estimated population by county 2015
summary(cancer$popEst2015)
hist(cancer$popEst2015, breaks = 100, main = "Estimated population by county 2015", xlab = NULL)
# There are 42 countries with population over 1 million
over1m = subset(cancer, popEst2015 >= 1000000)
nrow(over1m)
hist(over1m$popEst2015, breaks = 20, main = "Estimated population by county 2015", xlab = NULL)
# One county has over 10 million
over10m = subset(cancer, popEst2015 >= 10000000)
nrow(over10m)
```

```{r}
# Median Age
summary(cancer$MedianAge)
hist(cancer$MedianAge, breaks = 10, main = "Median Age", xlab = NULL)

# There are 30 countries with the median age over 100
medianAgeOver100 = subset(cancer, MedianAge >= 100)
nrow(medianAgeOver100)
```

## Analysis of Key Relationships

```{r}
# Focus on the variables that exihibited stronger relationships with deathRate

## Top 3 Positive Correlations

cor(cancer$deathRate, cancer$povertyPercent)
cor(cancer$deathRate, cancer$PctHS25_Over)
cor(cancer$deathRate, cancer$PctPublicCoverage)

## Top 3 Negative Correlations
cor(cancer$deathRate, cancer$PctBachDeg25_Over)
cor(cancer$deathRate, cancer$PctPrivateCoverage)
cor(cancer$deathRate, cancer$PctMarriedHouseholds)

# Box Plots
boxplot(deathRate ~ povertyPercent, data = canc,
        main = "Death Rate",
  xlab = "Death Rate", ylab = "Poverty Percent")

hist(cancer$povertyPercent, breaks = 20, main = "Poverty Percent", xlab = NULL)

boxplot(deathRate ~ PctBachDeg25_Over, data = canc,
        main = "Death Rate",
  xlab = "Death Rate", ylab = "PctBachDeg25_Over")

hist(cancer$PctBachDeg25_Over, breaks = 20, main = "PctBachDeg25_Over", xlab = NULL)
```

## Objective

# Perform an exploratory analysis to understand how county-level characteristics are related to cancer mortality.