---
title: "Cancer Mortality Exploration"
author: "Andrew Carlson, Brandon Cummings, Tako Hisada"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Research Question

Our team was hired by a health government agency that would like to understand factors that predict cancer mortality rates. Their ultimate goal is to identify communities for social interventions and of understanding which interventions are likely to have the most impact. Our main objective is to perform an exploratory analysis to understand how county-level characteristics are related to cancer mortality.

# Dataset Analysis

```{r load}
Cancer = read.csv('cancer.csv')  
```

This dataset consists of `r length(colnames(Cancer)) - 1` variables (not including the index column), all pertaining to county level information. Overall there were about `r nrow(Cancer)` observations per variable.  

The types of variables present in the dataset can be categorized into 8 groups:  
  
1) Region  
2) Population  
3) Birthrate  
4) Race  
5) Marital Status  
6) Insurance coverage  
7) Income status  
8) Education


All variables in dataset:

```{r}
colnames(Cancer)  
```
  
\newpage
# Data Quality

Overall the data quality was reasonable and usable. There were some observations in different decimal states, many NAs, and some variables that didn't seem relevant to cancer mortality at all. Other than that we found the data to be easy to analyze. Below are some data observations and assumptions:

##### "deathRate" - This is the column that we have assumed is the number of average yearly deaths per county.  

##### "MedianAge" - This variable is the median age for a county, the dataset column had a range of 22-624, when analyzing this correlation we trimmed all numbers above 65 due to the numbers after 65 started in the 300s.  

##### "PctSomeCol18_2"4 - This is the percent of some college attended between the age of 18-24. This column only had 762 of 3047 observations that were not NA. We still used this column when analyzing correlation, but it is worth noting that we removed all NAs.   

##### "Race" - When it came to the percentage of race for each county, we noticed that a mojority of the counties surveyed were "white". This may or may not be a significant datapoint, but it may lead to assumptions about populations that are incorrect.  

##### "avgAnnCount" - This was clarified as "2009-2013 mean incidences per county", we did not know what "incidences" this was referring to, we ended up not finding a direct correlation with other important variables, so we did not make any further assumptions and left it out of our analysis.  

##### "AvgHouseholdSize" - This had 61 entries with less than 1, meaning that there are observations of 0 or negative household sizes, we removed these when analyzing houshold size with other key variables.  

##### "PctEmployed16_Over" - There were 152 missing observations in this column, we removed these NAs from our dataset when analyzing this column with other key variables.

\pagebreak


# Analysis of Key Variables and Relationships


```{r include=FALSE}
summary(Cancer)
attach(Cancer)
par(mar=c(20,4.1,4.1,2.1))
```

```{r}
# convenient wrapper function for a prettier histogram
histWithMean <- function(vec, name) {
  hist(vec, col=rgb(0,0,1,1/4), main=paste("Histogram of ", name), xlab=name)
  # add a red line down the mean
  abline(v = mean(vec, na.rm=TRUE), col="red", lwd=2)
}
```

The dependant variable for this analysis is `deathRate`, which is assumed to be the death rate from cancer.

```{r}
histWithMean(deathRate, "deathRate")
```

\pagebreak

# Unclean Data

```{r}
# function that counts the number of elements in a vector that satisfy the predicate
# convenient for checking certain sanity bounds and counting how many are out of the bounds
count.by <- function(vec, predicate) {
  yes <- 0
  no <- 0
  for (n in vec) {
    if (predicate(n)) {
      yes <- yes + 1
    } else {
      no <- no + 1
    }
  }
  return(c(yes, no))
}
```


61 of the `AvgHouseholdSize` entries are less than 1. This is probably a coding error. A mean less than 1 for a set of integers is only possible if some values are 0 or negative. These values are nonsensical for a household size.

```{r}
count.by(AvgHouseholdSize, function(num) num < 1)
```


30 of the `MedianAge` entries are greater than 200. This seems flagrantly improbable.

```{r}
count.by(MedianAge, function(num) num > 200)
```


152 of the `PctEmployed16_Over` entries are `NA`.

```{r}
count.by(PctEmployed16_Over, is.na)
```


2285 of the `PctSomeCol18_24` entries are `NA`.

```{r}
count.by(PctSomeCol18_24, is.na)
```

\pagebreak

# Correlated Variables

Here are some histograms of the variables that turned out to be related to `deathRate`.

`medIncome` looks like a positively skewed distribution. In fact, in some populations it may look more like a power law distribution than a normal [[link]](https://arxiv.org/pdf/1304.0212.pdf). If we plot `log(medIncome)`, it *looks* closer to a normal distribution. We can check this transformation for correlation with `deathRate` in addition to the plain `medIncome` variable.

```{r,echo=FALSE, out.width='.49\\linewidth', fig.width=4, fig.height=4, fig.show='hold', fig.align='center'}
histWithMean(medIncome, "medIncome")
histWithMean(log(medIncome), "log(medIncome)")
```

```{r}
Cancer$logMedIncome <- log(medIncome)
```

The rest look like clean, valid data. There are no obvious transformations to apply.

```{r,echo=FALSE, out.width='.49\\linewidth', fig.width=4, fig.height=4, fig.show='hold', fig.align='center'}
histWithMean(PctBachDeg25_Over, "PctBachDeg25_Over")
histWithMean(povertyPercent, "povertyPercent")
histWithMean(PctEmployed16_Over, "PctEmployed16_Over")
histWithMean(PctHS25_Over, "PctHS25_Over")
histWithMean(PctPublicCoverage, "PctPublicCoverage")
histWithMean(PctPrivateCoverage, "PctPrivateCoverage")
histWithMean(PctUnemployed16_Over, "PctUnemployed16_Over")
```

```{r include=FALSE}
# some columns were added, so lets update the attached version
detach(Cancer)
attach(Cancer)
```

\pagebreak

# Finding strongest correlations

The numeric variables were taken. The correlation with each numeric variable was calculated and sorted by descending absolute value.

```{r}
# get just the numeric columns
numericColumns <- sapply(Cancer, is.numeric)
NumericCancer <- Cancer[, numericColumns]
# get each correlations with each column
correlations <- apply(NumericCancer, 2, function(col) cor(col, deathRate))
correlations <- correlations[!is.na(correlations)]
```

Now we have a vector of all the correlations. We just filtered out the `NA`s, which includes `PctEmployed16_Over` because some of the entries were `NA`. We'll have to add it back manually after dealing with the `NA`s.

```{r}
# clean the NAs out of PctEmployed16_Over and calculate correlation
cleanPctEmployed16_Over <- !is.na(PctEmployed16_Over)
corPctEmployed16_Over <- cor(PctEmployed16_Over[cleanPctEmployed16_Over], deathRate[cleanPctEmployed16_Over])
# append it to the vector of correlations and name the entry
correlations <- c(correlations, corPctEmployed16_Over)
names(correlations)[length(correlations)] <- "PctEmployed16_Over"
```

Now we can focus on the correlations that are significant.

```{r}
# sort them
correlations <- correlations[order(abs(correlations), decreasing=TRUE)]
correlations <- correlations[2:length(correlations)]
correlations <- correlations[abs(correlations) >= 0.3]
correlations
```

\pagebreak

We will consider correlations of `0.3` or greater a significant association. This includes 9 of the variables, one of which is our transformed log(medianIncome). This actually had stronger correlation with `deathRate` than `medIncome`.

# Plot deathRate with all variables with at least a weak correlation

```{r,echo=FALSE, out.width='.49\\linewidth', fig.width=4, fig.height=4, fig.show='hold', fig.align='center'}

#par(mar=c(8,4,6,2))
plotInfo <- function(vec, name, corr) {
  plot(vec, deathRate, cex=0.2, sub=paste("r = ", corr), xlab=name, main=paste("deathRate vs ", name))
}

for (i in 1:length(correlations)) {
  name <- names(correlations)[i]
  plotInfo(unlist(Cancer[name]), name, correlations[i])
}
```

# Analysis of Secondary Effects

## Racial Groups

We identified certain racial groups are exhitibing higher cancer death rate than others, namely black people (`PctBlack`).

```{r,echo=FALSE, out.width='.49\\linewidth', fig.width=4, fig.height=4, fig.show='hold', fig.align='center'}

# Race-related variables
raceVars = Cancer[, c("PctWhite", "PctBlack", "PctAsian", "PctOtherRace")]

# get each correlations with each column
raceCors <- apply(raceVars, 2, function(col) cor(col, deathRate))

for (i in 1:length(raceCors)) {
  name <- names(raceCors)[i]
  plotInfo(unlist(Cancer[name]), name, raceCors[i])
}
```

While the other groups are showing negative correlations with the death rate, `PctBlack` is showing a mildly strong correlation of 0.257. This at the first glance makes it seem as though black people are somehow more susceptible to cancer.

Looking further, we find `PctBlack` has strong positive correlations with `povertyPercent` and `PctUnemployed16_Over`, which are amongst the variables we have identified to have the strongest positive correlations with `deathRate`. 

```{r,echo=FALSE, out.width='.49\\linewidth', fig.width=4, fig.height=4, fig.show='hold', fig.align='center'}

plotInfoPctBlack <- function(vec, name, corr) {
  plot(vec, Cancer$PctBlack, cex=0.2, sub=paste("r = ", corr), xlab=name, main=paste("PctBlack vs ", name))
}

plotInfoPctBlack(Cancer$povertyPercent, "povertyPercent", cor(Cancer$PctBlack, Cancer$povertyPercent))
plotInfoPctBlack(Cancer$PctUnemployed16_Over, "PctUnemployed16_Over", cor(Cancer$PctBlack, Cancer$PctUnemployed16_Over))
```

We also see negative correlations between `PctBlack` and the variables we have identified to have strong negative correlations with `deathRate` such as `logMedIncome`, `PctBachDeg25_Over` and `PctPrivateCoverage`.

```{r,echo=FALSE, out.width='.49\\linewidth', fig.width=4, fig.height=4, fig.show='hold', fig.align='center'}
plotInfoPctBlack(Cancer$logMedIncome, "logMedIncome", cor(Cancer$PctBlack, Cancer$logMedIncome))
plotInfoPctBlack(Cancer$PctBachDeg25_Over, "PctBachDeg25_Over", cor(Cancer$PctBlack, Cancer$PctBachDeg25_Over))
plotInfoPctBlack(Cancer$PctPrivateCoverage, "PctPrivateCoverage", cor(Cancer$PctBlack, Cancer$PctPrivateCoverage))
```

Therefore we believe it is not their race that is increasing the cancer death rate for black people but rather the group's relatively poor econoimc standing that is contributing to the elevated `deathRate`. 

## Martial-Status

We are alsso seeing relatively strong negative correlations between martial status-releated variables `PercentMarried` and `PctMarriedHouseholds` and `deathRate`. Marriage somehow decreases the chance of dying from cancer. Similarly to racial groups, however, after looking at the correlations between the martial-status variables and the same set of variables strongly correlated with `deathRate`, we see married people are considerably less likely to be in poverty. This explains the negative correlation observed between the martial status-related variables and `deathRate` rather than people's martial statuses themselves somehow magically reducing the cancer death rate. 

```{r,echo=FALSE, out.width='.49\\linewidth', fig.width=4, fig.height=4, fig.show='hold', fig.align='center'}

# Martial status-related variables
marriedVars = Cancer[, c("PercentMarried", "PctMarriedHouseholds")]

# get each correlations with each column
marriedCors <- apply(marriedVars, 2, function(col) cor(col, deathRate))

for (i in 1:length(marriedCors)) {
  name <- names(marriedCors)[i]
  plotInfo(unlist(Cancer[name]), name, marriedCors[i])
}
```

```{r,echo=FALSE, out.width='.49\\linewidth', fig.width=4, fig.height=4, fig.show='hold', fig.align='center'}

# `PctMarried`

plotInfoPctMarried <- function(vec, name, corr) {
  plot(vec, Cancer$PercentMarried, cex=0.2, sub=paste("r = ", corr), xlab=name, main=paste("PercentMarried vs ", name))
}

## Variables positively correlated with `deathRate` 
plotInfoPctMarried(Cancer$povertyPercent, "povertyPercent", cor(Cancer$PercentMarried, Cancer$povertyPercent))
plotInfoPctMarried(Cancer$PctUnemployed16_Over, "PctUnemployed16_Over", cor(Cancer$PercentMarried, Cancer$PctUnemployed16_Over))

## Variables negetively correlated with `deathRate` 
plotInfoPctMarried(Cancer$logMedIncome, "logMedIncome", cor(Cancer$PercentMarried, Cancer$logMedIncome))
plotInfoPctMarried(Cancer$PctBachDeg25_Over, "PctBachDeg25_Over", cor(Cancer$PercentMarried, Cancer$PctBachDeg25_Over))
plotInfoPctMarried(Cancer$PctPrivateCoverage, "PctPrivateCoverage", cor(Cancer$PercentMarried, Cancer$PctPrivateCoverage))

# `PctMarriedHouseholds`

plotInfoPctMarriedHouseholds <- function(vec, name, corr) {
  plot(vec, Cancer$PctMarriedHouseholds, cex=0.2, sub=paste("r = ", corr), xlab=name, main=paste("PctMarriedHouseholds vs ", name))
}

## Variables positively correlated with `deathRate` 
plotInfoPctMarriedHouseholds(Cancer$povertyPercent, "povertyPercent", cor(Cancer$PctMarriedHouseholds, Cancer$povertyPercent))
plotInfoPctMarriedHouseholds(Cancer$PctUnemployed16_Over, "PctUnemployed16_Over", cor(Cancer$PctMarriedHouseholds, Cancer$PctUnemployed16_Over))

## Variables negetively correlated with `deathRate` 
plotInfoPctMarriedHouseholds(Cancer$logMedIncome, "logMedIncome", cor(Cancer$PctMarriedHouseholds, Cancer$logMedIncome))
plotInfoPctMarriedHouseholds(Cancer$PctBachDeg25_Over, "PctBachDeg25_Over", cor(Cancer$PctMarriedHouseholds, Cancer$PctBachDeg25_Over))
plotInfoPctMarriedHouseholds(Cancer$PctPrivateCoverage, "PctPrivateCoverage", cor(Cancer$PctMarriedHouseholds, Cancer$PctPrivateCoverage))
```

# Conclusion

Of the 30 variables included in the Cancer Mortality dataset, we are observing strongest correlations between `deathRate` and the below 9 variables:

```{r,echo=FALSE, out.width='.49\\linewidth', fig.width=4, fig.height=4, fig.show='hold', fig.align='center'}
names(correlations)
```

These variables fall under the categories Insurance coverage, Income Status and Education while variables that are categorized under Region, Population, Birthrate, Race and Martial Status are not showing as strong of relationships with `deathRate`.

From this, we are able to conclude that for the US counties included in the datset, people's economic standing is the deciding factor for their cancer mortality rate rather than geographical factors such as region and population. Other factors such as race and martial status also appear to be directly correlated with `deathRate`. However, we found that these variables have strong correlations with Income Status and Insurance Coverage variables which lead us to suspect it is not the racial demographics or martial statuses of the residents that are diretly influencing `deathRate` but rather their economic implications that are affecting the cancer death rate in these counties.